---
title: "FinalProject"
author: "Devin Domeyer"
date: "4/11/2022"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(dplyr)
library(viridis)
library(cowplot)
library(sf)
library(leaflet)
library(mapview)
library(knitr)
library(scales)
library(lme4)
library(lmerTest)
library(agricolae)

setwd("~/Desktop/Duke/Data Analytics/Domeyer_ENV872_EDA_FinalProject")

EPA_O3_2018 <- read.csv("./Data/Raw/EPAair_O3_NC2018_raw_copy.csv", stringsAsFactors = TRUE)
EPA_O3_2019 <- read.csv("./Data/Raw/EPAair_O3_NC2019_raw_copy.csv", stringsAsFactors = TRUE)
EPA_PM25_2018 <- read.csv("./Data/Raw/EPAair_PM25_NC2018_raw_copy.csv", stringsAsFactors = TRUE)
EPA_PM25_2019 <- read.csv("./Data/Raw/EPAair_PM25_NC2019_raw_copy.csv", stringsAsFactors = TRUE)

mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right",
        legend.title = element_text(size=10),
        legend.text = element_text(size=8),
        plot.title = element_text(hjust = 0.5))
        
theme_set(mytheme)
```

# Research Questions

Question 1: Is the air quality index for Ozone and PM2.5 significantly worse in the summer (June - August) compared to the rest of the year?
Question 2: What county has the highest average Ozone and PM2.5 levels?

# Data Wrangling

```{r wrangle 1}
EPA_O3_2018$Date <- mdy(EPA_O3_2018$Date)
EPA_O3_2019$Date <- mdy(EPA_O3_2019$Date)
EPA_PM25_2018$Date <- mdy(EPA_PM25_2018$Date)
EPA_PM25_2019$Date <- mdy(EPA_PM25_2019$Date)

EPA_O3_2018_subset <- select(EPA_O3_2018, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, SITE_LATITUDE, COUNTY:SITE_LONGITUDE)
EPA_O3_2019_subset <- select(EPA_O3_2019, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, SITE_LATITUDE, COUNTY:SITE_LONGITUDE)
EPA_PM25_2018_subset <- select(EPA_PM25_2018, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, SITE_LATITUDE, COUNTY, SITE_LONGITUDE)
EPA_PM25_2019_subset <- select(EPA_PM25_2019, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, SITE_LATITUDE, COUNTY, SITE_LONGITUDE)

EPA_PM25_2018_subset$AQS_PARAMETER_DESC <- "PM2.5"
EPA_PM25_2019_subset$AQS_PARAMETER_DESC <- "PM2.5"

write.csv(EPA_O3_2018_subset, row.names = FALSE, file = "./Data/Processed/EPAair_O3_2018_Processed.csv")
write.csv(EPA_O3_2019_subset, row.names = FALSE, file = "./Data/Processed/EPAair_O3_2019_Processed.csv")
write.csv(EPA_PM25_2019_subset, row.names = FALSE, file = "./Data/Processed/EPAair_PM25_2019_Processed.csv")
write.csv(EPA_PM25_2018_subset, row.names = FALSE, file = "./Data/Processed/EPAair_PM25_2018_Processed.csv")
```

```{r wrangle 2}
# Find counties common across all four dataframes
O3_Counties <- intersect(EPA_O3_2018_subset$COUNTY, EPA_O3_2019_subset$COUNTY)
PM2.5_Counties <- intersect(EPA_PM25_2018_subset$COUNTY, EPA_PM25_2019_subset$COUNTY)
Common_Counties <- intersect(O3_Counties, PM2.5_Counties)

# Combine all dataframes
EPA_data <- rbind(EPA_O3_2018_subset, EPA_O3_2019_subset, EPA_PM25_2018_subset, EPA_PM25_2019_subset)

EPA_data_filtered <- EPA_data %>%
  filter(COUNTY %in% c("Avery", "Buncombe", "Cumberland", "Durham", "Edgecombe", "Forsyth", "Guilford", "Haywood", "Johnston", "Mecklenburg", "Montgomery", "New Hanover", "Pitt", "Swain", "Wake")) %>%
  mutate(month = month(Date)) %>%
  mutate(year = year(Date)) %>%
  group_by(month, AQS_PARAMETER_DESC, COUNTY) %>%
  summarise(meanAQI = mean(DAILY_AQI_VALUE)) %>%
  pivot_wider(names_from = "AQS_PARAMETER_DESC", values_from = "meanAQI")

write.csv(EPA_data_filtered, row.names = FALSE, file = "./Data/Processed/EPAair_PM25_O3_filtered.csv")
```
# Data Information

```{r information}
colnames(EPA_data_filtered)[1] <- "Month"
colnames(EPA_data_filtered)[2] <- "County"
County_summary <- EPA_data_filtered %>%
  group_by(County) %>%
  summarise(Mean_Ozone = mean(Ozone, na.rm=T),
            SD_Ozone = sd(Ozone, na.rm=T),
            Mean_PM2.5 = mean(PM2.5, na.rm=T),
            SD_PM2.5 = sd(PM2.5, na.rm=T))

Month_summary <- EPA_data_filtered %>%
  group_by(Month) %>%
  summarise(Mean_Ozone = mean(Ozone, na.rm=T),
            SD_Ozone = sd(Ozone, na.rm=T),
            Mean_PM2.5 = mean(PM2.5, na.rm=T),
            SD_PM2.5 = sd(PM2.5, na.rm=T))

knitr::kable(County_summary, caption ="County summary of pollutants (ppm) in North Carolina, 2018 - 2019")
knitr::kable(Month_summary, caption ="Monthly summary of pollutants (ppm) in North Carolina, 2018 - 2019")
            
```

Detail                    | Description
------------------------- | -----------------                   
Data Source               | EPA Outdoor Air Quality Data
Variables Used            | Month, Air Quality Index (AQS), County, Latitude, Longitude
Date Range                | 2018 - 2019

# Data Exploration

```{r exploration}
# Spatial extent of dataset
counties_sf <- st_read("./Data/Raw/cb_2018_us_county_20m.shp") %>% 
  filter(STATEFP == 37)

select_sf <- counties_sf %>% 
  filter(STATEFP == 37) %>%
  filter(NAME %in% c("Avery", "Buncombe", "Cumberland", "Durham", "Edgecombe", "Forsyth", "Guilford", "Haywood", "Johnston", "Mecklenburg", "Montgomery", "New Hanover", "Pitt", "Swain", "Wake"))

ggplot() +
  geom_sf(data = counties_sf, color = 'grey') +
  geom_sf(data = select_sf, aes(fill = NAME)) +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(), 
        axis.ticks.y=element_blank(),
        axis.line=element_blank()) +
  labs(fill="County")
  
# Graph representation of ozone and pm2.5
ozone_plot <- ggplot(EPA_data_filtered, aes(x=Month, y=Ozone)) +
  geom_line(aes(color=County)) +
  scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12)) + 
  ylab("Mean Ozone (ppm)")

print(ozone_plot)

pm25_plot <- ggplot(EPA_data_filtered, aes(x=Month, y=PM2.5)) +
  geom_line(aes(color=County)) +
  scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12)) + 
  ylab("Mean PM2.5 (ppm)")

print(pm25_plot)

```

# Analysis
```{r analysis question 1}
lm1 <- lm(data=EPA_data_filtered, Ozone ~ County + Month)
step(lm1)

# Best fit includes Month as the only factor.
lm2 <- lm(data=EPA_data_filtered, Ozone ~ Month)
summary(lm2)
month.groups <- HSD.test(lm2, "Month", group = TRUE)
month.groups

# Experiment by adding County as random effect. Test confirms it does not improve the model.
lm2.1 <- lmer(data=EPA_data_filtered, Ozone ~ Month + (1|County))
summary(lm2.1)
step(lm2.1)

lm3 <- lm(data=EPA_data_filtered, PM2.5 ~ County + Month)
step(lm3)

# Best fit includes County as the only factor. 
lm4 <- lm(data=EPA_data_filtered, PM2.5 ~ County)
summary(lm4)

county.groups <- HSD.test(lm4, "County", group = TRUE)
county.groups

```

# Results

```{r results}
# Monthly plot
colors <- c("Ozone" = "blue", "PM2.5" = "orange")

month_plot <- ggplot(Month_summary) +
  geom_line(aes(x=Month, y=Mean_Ozone, color = "Ozone"), size=2) +
  geom_line(aes(x=Month, y=Mean_PM2.5, color = "PM2.5"), size=2) +
  scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12)) + 
  labs(x = "Month",
       y = "Mean Pollutant Conc. (ppm)",
       color = "Legend") +
  scale_color_manual(values=colors)

print(month_plot)

# County Plot
county_plot <- ggplot(EPA_data_filtered) +
  geom_boxplot(aes(x=reorder(County,PM2.5,na.rm=TRUE), y=PM2.5)) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

print(county_plot)

```
